Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-07-02 manuscripts/reply/reply-main.tex
Working on reply ms.
Moving coin-flipping example to supporting information.

This complicates the meaning of model posterior probabilities when Bayes' rule
is violated by informing priors with the same data to be analyzed.
For empirical Bayesian parameter estimation, the objective is that by giving the
data more weight relative to the prior, the posterior distribution will be
peaked near the true value(s) of the model's parameter(s).
There is no such justification for empirical Bayesian model choice, because
there are no true values for the model probabilities being estimated.
By using the data twice, we fail to account for prior uncertainty and mislead
our posterior beliefs in the models being compared; we will be over confident
in some models and under confident in others.
Nonetheless, empirical Bayesian model choice does perform well for some
problems.
Particularly, in cases where large aggregate data sets are used for many
parallel model-choice problems simultaneously, pooling information to inform
priors can lead to favorable group-wise frequentist coverage across tests
\citep{Efron2008}.
However, this is far removed from the single model-choice problem of \msb.

\begin{linenomath}
These distinctions between Bayesian parameter estimation and model choice can
be illustrated with a simple example.
Let us say we are interested in the fairness of a particular coin, and we
denote the unknown probability of it landing heads as \myTheta{}.
More specifically, we are interested in the probability of two models, $M_1$
and $M_2$.
In both models the outcomes of flipping the coin are assumed to be binomially
distributed, but under $M_1$ the coin is weighted toward landing heads (i.e.,
$\myTheta{} > 0.5)$), whereas under $M_2$, the coin is weighted toward landing
tails (i.e., $\myTheta{} < 0.5$).
We already have data from flipping a different coin 20 times that landed both
heads and tails 10 times each, and so we decide to use these data in specifying
a beta prior on fairness of the new coin of $beta(a=10, b=10)$
(Figure~S\ref{figCoinFlip}).
We collect data by flipping the coin of interest $N=10$ times, $y=3$ of which
land heads.
Given the beta distribution is a conjugate prior for a binomial likelihood, the
posterior distribution has the nice analytical form $\theta \given y,N \sim
beta(a + y, b + N - y)$, which for the new dataset is simply $beta(13, 17)$
(Figure~S\ref{figCoinFlip}).
The maximum a posteriori (MAP) estimate of the probability of heads is 0.429,
and following Equation~\ref{eq:marginallikelhiood} the marginal likelihoods of
our models of interest are
\begin{equation}
    p(y=3, N=10 \given M_1) = \int_{0.5}^{1} p(y=3, N=10 \given
    \myTheta{}, M_1)p(\myTheta{} \given M_1) d\myTheta{} \approx 0.029,
\end{equation}
and
\begin{equation}
    p(y=3, N=10 \given M_2) = \int_{0}^{0.5} p(y=3, N=10 \given
    \myTheta{}, M_2)p(\myTheta{} \given M_2) d\myTheta{} \approx 0.097.
\end{equation}
Given the models have equal probability under our prior, we can calculate the
posterior probability of Model 1 as
\begin{equation}
    p(M_1 \given y=3, N=10) = \frac{p(y=3, N=10 \given M_1)}{p(y=3, N=10 \given
    M_1) + p(y=3, N=10 \given M_2)} \approx 0.23.
\end{equation}
This is the correct posterior probability of Model 1 given our prior
and data.
\end{linenomath}

To give the data more weight relative to the prior, we could use it twice, and
calculate an empirical Bayes estimate using a prior of $beta(13,17)$.  This
results in a ``posterior'' distribution of $beta(16, 24)$
(Figure~S\ref{figCoinFlip}), with a MAP estimate of 0.395, and $p(M_1 \given
y=3, N=10) = 0.10$.
The estimated posterior distribution of the parameter, and resulting MAP
estimate, is similar whether or not an empirically informed prior is used.
However, the posterior probability of Model 1 is very sensitive to the
empirical prior, decreasing by 56\%.
By using the empirically informed prior, we ignored prior uncertainty, leading
to an underestimate of our posterior uncertainty (Figure~S\ref{figCoinFlip}).
While this did not greatly affect our estimate of \myTheta{}, it misled us
to be overconfident in Model 2.
% Our mode estimate of \myTheta{} under the empirical prior might be closer
% to the true value than our posterior mode.
% Let us assume the truth of the matter is that the coin under study has a broad
% flat edge, and as a result lands on its edge at a certain frequency.
% Hence, both of our models are incorrect (i.e., the outcomes of flipping the
% coin are not binomially distributed).

% This can be demonstrated with a simple, albeit contrived, example.
% Let us say that principal investigator Mary and her new postdoc Will are
% interested in the hypothesis that the mass of George Washington's periwig
% renders the quarter dollar of the United States unfair.
% That is to say their null hypothesis is that the probability of a US
% quarter landing heads when tossed is less than 0.5 ($\theta < 0.5$).
% They can certainly evaluate this hypothesis, they can have undergraduate
% worker, Joe, flip the coin for them while they tabulate the results.
% But being Bayesians, before they call Joe into the lab, they agree on a
% prior probability to place on the set of all possible probabilities that
% the quarter will land heads when it is flipped.
% Given that neither of them have a quarter, and their prior knowledge that Joe
% moonlights as a magician, and is notorious for performing coin and card tricks
% in the lab, they suspect there is a good chance that the quarter Joe uses will
% be either two-headed or two-tailed.
% So, knowing that the beta distribution is the conjugate prior for a binomial
% likelihood, they decide to use a $beta(a=0.5, b=0.5)$ prior distribution
% (Figure~\ref{figCoinFlip}).

% Mary calls Joe into the lab, confirms that he has a quarter, and tells him to
% begin flipping it.
% After five tosses, four of which was heads, Joe decides that academics are
% crazy and leaves the lab to pursue a major in theatre.  Mary and Will, both
% being computational biologists, are satisfied with their
% empirical dataset of $y = 4$ heads out of $N = 5$ trials.
% They know from the conjugacy of the beta prior, that the posterior distribution
% has the nice analytical form $\theta|y,N \sim beta(a + y, b + N - y)$, which
% in this case is simply $beta(4.5, 1.5)$; this is the true posterior distribution
% of $\theta$ given their prior belief and data (Figure~\ref{figCoinFlip}).
% This allows them to plug these values into the beta cumulative distribution
% function to determine that the posterior probability of their hypothesis is
% $p(\theta < 0.5 | y=1, N=5) = 0.088$.
% Given their prior belief and dataset, this indeed is the correct posterior
% probability of the hypothesis, and Mary and Will should now update their
% posterior belief accordingly.

% However, reflecting upon the results of their experiment
% (Figure~\ref{figCoinFlip}), Mary and Will regret their choice of prior.
% Their prior looks very ``poorly selected,'' and if they had only known that the
% coin was fair before the data were collected, they would have selected a much
% better prior.
% Clearly, from their results, they should have used a prior centered
% around 0.2; their data suggest a prior of $beta(4.5, 1.5)$ would have been
% much ``better.''
% Rather than resort to flipping the coin five more times, Mary and Will decide
% to redo their analysis using the much better ``prior'' of $p(\theta) \sim
% beta(4.5, 1.5)$.
% This gives a ``posterior'' of $\theta|y,N \sim beta(8.5, 2.5)$, and a
% probability of their hypothesis of $p(\theta < 0.5 | y=1, N=5) = 0.026$.
% Now convinced that the US quarter is unfair, albeit not due to the mass of
% President Washington's head as they hypothesized, Will begins composing an
% e-mail of complaint addressed to the U.S.\ Mint.

% If Joe's flips were a representative sample, Mary and Will's empirical Bayes
% estimate might very well be a better point estimate for the parameter
% \myTheta{}.
% However, their empirical Bayes estimate of the probability of their hypothesis
% is incorrect and biased.
% This simple example shows how parameter estimation is fundamentally different
% from estimating the probability of a model.
% While empirically informed priors can be used to obtain well-behaved parameter
% estimators, using them for model choice is much less certain.

>>>>

This complicates the meaning of model posterior probabilities when Bayes' rule
is violated by informing priors with the same data to be analyzed.
For empirical Bayesian parameter estimation, the objective is that by giving the
data more weight relative to the prior, the posterior distribution will be
peaked near the true value(s) of the model's parameter(s).
There is no such justification for empirical Bayesian model choice, because
there are no true values for the model probabilities being estimated.
By using the data twice, we fail to account for prior uncertainty and mislead
our posterior beliefs in the models being compared; we will be over confident
in some models and under confident in others.
Nonetheless, empirical Bayesian model choice does perform well for some
problems.
Particularly, in cases where large aggregate data sets are used for many
parallel model-choice problems simultaneously, pooling information to inform
priors can lead to favorable group-wise frequentist coverage across tests
\citep{Efron2008}.
However, this is far removed from the single model-choice problem of \msb.
In the Supporting Information we use a simple example to help highlight the
distinctions between Bayesian parameter estimation and model choice.

### 2014-07-02 manuscripts/reply/reply-si-main.tex
Working on reply ms.
Moving coin-flipping example to supporting information.

>>>>

\section{Theoretical implications of empirical priors for Bayesian model
choice---A simple example}
\begin{linenomath}
The distinctions between Bayesian parameter estimation and model choice
discussed in the main text can be illustrated with a simple example.
Let us say we are interested in the fairness of a particular coin, and we
denote the unknown probability of it landing heads as \myTheta{}.
More specifically, we are interested in the probability of two models, $M_1$
and $M_2$.
In both models the outcomes of flipping the coin are assumed to be binomially
distributed, but under $M_1$ the coin is weighted toward landing heads (i.e.,
$\myTheta{} > 0.5)$), whereas under $M_2$, the coin is weighted toward landing
tails (i.e., $\myTheta{} < 0.5$).
We already have data from flipping a different coin 20 times that landed both
heads and tails 10 times each, and so we decide to use these data in specifying
a beta prior on fairness of the new coin of $beta(a=10, b=10)$
(Figure~S\ref{figCoinFlip}).
We collect data by flipping the coin of interest $N=10$ times, $y=3$ of which
land heads.
Given the beta distribution is a conjugate prior for a binomial likelihood, the
posterior distribution has the nice analytical form $\theta \given y,N \sim
beta(a + y, b + N - y)$, which for the new dataset is simply $beta(13, 17)$
(Figure~S\ref{figCoinFlip}).
The maximum a posteriori (MAP) estimate of the probability of heads is 0.429,
and following Equation~\ref{eq:marginallikelhiood} in the main text the
marginal likelihoods of our models of interest are
\begin{equation}
    p(y=3, N=10 \given M_1) = \int_{0.5}^{1} p(y=3, N=10 \given
    \myTheta{}, M_1)p(\myTheta{} \given M_1) d\myTheta{} \approx 0.029,
\end{equation}
and
\begin{equation}
    p(y=3, N=10 \given M_2) = \int_{0}^{0.5} p(y=3, N=10 \given
    \myTheta{}, M_2)p(\myTheta{} \given M_2) d\myTheta{} \approx 0.097.
\end{equation}
Given the models have equal probability under our prior, we can calculate the
posterior probability of Model 1 as
\begin{equation}
    p(M_1 \given y=3, N=10) = \frac{p(y=3, N=10 \given M_1)}{p(y=3, N=10 \given
    M_1) + p(y=3, N=10 \given M_2)} \approx 0.23.
\end{equation}
This is the correct posterior probability of Model 1 given our prior
and data.
\end{linenomath}

To give the data more weight relative to the prior, we could use it twice, and
calculate an empirical Bayes estimate using a prior of $beta(13,17)$.  This
results in a ``posterior'' distribution of $beta(16, 24)$
(Figure~S\ref{figCoinFlip}), with a MAP estimate of 0.395, and $p(M_1 \given
y=3, N=10) = 0.10$.
The estimated posterior distribution of the parameter, and resulting MAP
estimate, is similar whether or not an empirically informed prior is used.
However, the posterior probability of Model 1 is very sensitive to the
empirical prior, decreasing by 56\%.
By using the empirically informed prior, we ignored prior uncertainty, leading
to an underestimate of our posterior uncertainty (Figure~S\ref{figCoinFlip}).
While this did not greatly affect our estimate of \myTheta{}, it misled us
to be overconfident in Model 2.
% Our mode estimate of \myTheta{} under the empirical prior might be closer
% to the true value than our posterior mode.
% Let us assume the truth of the matter is that the coin under study has a broad
% flat edge, and as a result lands on its edge at a certain frequency.
% Hence, both of our models are incorrect (i.e., the outcomes of flipping the
% coin are not binomially distributed).

% This can be demonstrated with a simple, albeit contrived, example.
% Let us say that principal investigator Mary and her new postdoc Will are
% interested in the hypothesis that the mass of George Washington's periwig
% renders the quarter dollar of the United States unfair.
% That is to say their null hypothesis is that the probability of a US
% quarter landing heads when tossed is less than 0.5 ($\theta < 0.5$).
% They can certainly evaluate this hypothesis, they can have undergraduate
% worker, Joe, flip the coin for them while they tabulate the results.
% But being Bayesians, before they call Joe into the lab, they agree on a
% prior probability to place on the set of all possible probabilities that
% the quarter will land heads when it is flipped.
% Given that neither of them have a quarter, and their prior knowledge that Joe
% moonlights as a magician, and is notorious for performing coin and card tricks
% in the lab, they suspect there is a good chance that the quarter Joe uses will
% be either two-headed or two-tailed.
% So, knowing that the beta distribution is the conjugate prior for a binomial
% likelihood, they decide to use a $beta(a=0.5, b=0.5)$ prior distribution
% (Figure~\ref{figCoinFlip}).

% Mary calls Joe into the lab, confirms that he has a quarter, and tells him to
% begin flipping it.
% After five tosses, four of which was heads, Joe decides that academics are
% crazy and leaves the lab to pursue a major in theatre.  Mary and Will, both
% being computational biologists, are satisfied with their
% empirical dataset of $y = 4$ heads out of $N = 5$ trials.
% They know from the conjugacy of the beta prior, that the posterior distribution
% has the nice analytical form $\theta|y,N \sim beta(a + y, b + N - y)$, which
% in this case is simply $beta(4.5, 1.5)$; this is the true posterior distribution
% of $\theta$ given their prior belief and data (Figure~\ref{figCoinFlip}).
% This allows them to plug these values into the beta cumulative distribution
% function to determine that the posterior probability of their hypothesis is
% $p(\theta < 0.5 | y=1, N=5) = 0.088$.
% Given their prior belief and dataset, this indeed is the correct posterior
% probability of the hypothesis, and Mary and Will should now update their
% posterior belief accordingly.

% However, reflecting upon the results of their experiment
% (Figure~\ref{figCoinFlip}), Mary and Will regret their choice of prior.
% Their prior looks very ``poorly selected,'' and if they had only known that the
% coin was fair before the data were collected, they would have selected a much
% better prior.
% Clearly, from their results, they should have used a prior centered
% around 0.2; their data suggest a prior of $beta(4.5, 1.5)$ would have been
% much ``better.''
% Rather than resort to flipping the coin five more times, Mary and Will decide
% to redo their analysis using the much better ``prior'' of $p(\theta) \sim
% beta(4.5, 1.5)$.
% This gives a ``posterior'' of $\theta|y,N \sim beta(8.5, 2.5)$, and a
% probability of their hypothesis of $p(\theta < 0.5 | y=1, N=5) = 0.026$.
% Now convinced that the US quarter is unfair, albeit not due to the mass of
% President Washington's head as they hypothesized, Will begins composing an
% e-mail of complaint addressed to the U.S.\ Mint.

% If Joe's flips were a representative sample, Mary and Will's empirical Bayes
% estimate might very well be a better point estimate for the parameter
% \myTheta{}.
% However, their empirical Bayes estimate of the probability of their hypothesis
% is incorrect and biased.
% This simple example shows how parameter estimation is fundamentally different
% from estimating the probability of a model.
% While empirically informed priors can be used to obtain well-behaved parameter
% estimators, using them for model choice is much less certain.
